Is Partner Modeling Mutual?

### 2014-08-27 mutual-modeling.tex
Texifying...
While here:
{project- ->}mutual-modeling.tex

@@

\usepackage{subcaption}

\usepackage[draft]{fixme}

@@

\newcommand{\M}[3]{{\mathcal{M}(#1, #2, #3)}}
\newcommand{\model}[3]{{$\mathcal{M}(#1, #2, #3)$}}
\newcommand{\refmodel}[2]{{$\mathcal{M}(#1, #2)$}}
\newcommand{\Model}[3]{{$\mathcal{M}^{\circ}(#1, #2, #3)$}}

@@

From its very beginning, CSCL research has been following Roschelle \& Teasley's
(1986) suggestion that collaborative learning has something to do with the
process of constructing and maintaining a shared understanding of the task at
hand. Building a shared/mutual understanding refers to the upper class of
collaborative learning situations, those in which students should build upon
each other's understanding to refine their own understanding. What is expected
to produce learning is not the mere fact that two students build the same
understanding but the cognitive effort they have to engage to build this shared
understanding (Schwartz, 1995). This effort can be observed by the frequency of
rich interactions, i.e. interactions whose occurrence has been related to
learning: (self-) explanations in cognitive science (Chi ; Webb), conflict
resolution in socio-cognitive theories (Doise), mutual regulation (Blaye,) in a
Vygostkian perspective, etc. The construction of a shared understanding has been
investigated for several years in psycholinguistics, under the  notion of
“grounding” (Clark \&Wilkes-Gibbs, 1986). However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned. The monitoring
and repair of mis-understanding explains for instance referential failures in
short dialogue episodes but does hardly predict conceptual change over longer
sessions (Dillenbourg \& Traum, 2006). The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:
"collaborative learning is associated with the increased cognitive-interactional
effort involved in the transition from learning to understand each other to
learning to understand the meanings of the semiotic tools that constitute the
mediators of interpersonal interaction" (Baker, Hansen, Joiner \& Traum, 1999).
Several scholars suggest that CSCL research should go deeper in the
understanding of how partners engage into shared meaning making (Stahl, 2006) or
'intersubjective' meaning making (Suthers (2006).

Paradoxically, while Clark's theory is somewhat too linguistic from a conceptual
change viewpoint, it is criticized at the same time as being too cognitivist by
some psycholinguists, i.e. as overestimating the amount shared knowledge and
mutual representations actually necessary to conduct a dialogue. The fundamental
issue, as old as philosophy, is the degree of coupling between the different
levels of dialogue, mostly between the lexical / syntactical level and the
deeper semantic levels. Pickering \& Garrod (2004) argue that the mutual
understanding starts mostly with a 'local alignment' at the level of the
linguistic representations, due to priming mechanisms, and that this superficial
alignment may – in some cases- lead to a 'global alignment' of the semantic
level.  For these authors, the convergence in dialogue, and even the repair of
some misunderstandings, is explained by this mimetic behavior more than by a
monitoring of each other's knowledge: "…interlocutors do not need to monitor and
develop full common ground as a regular, constant part of routine conversation,
as it would be unnecessary and far too costly. Establishment of full common
ground is, we argue, a specialized and non-automatic process that is used
primarily in times of difficulty (when radical misalignment becomes apparent)."
(p. 179). This view is actually not incompatible with Clark's grounding
criterion (REF): the degree of shared understanding that peers need to reach
depends upon the task they perform. For instance, a dialogue between two
surgeons might rely on superficial alignment if they talk about their friends
but has to guarantee accurate common grounds when talking about which
intervention will be conducted in which way on which patient.  Some CSCL
scholars stressed the 'illusion of shared understanding' (JACCO?), i.e. the
potential decoupling between linguistic alignment and actual shared meanings.

@@

Deep grounding or shared meaning making requires some cognitive load. For Clark
\& Wilkes-Gibbs (1986), what is important is not the individual effort made by
the receiver of a communicative act, but the overall least collaborative effort.
The cost of producing a perfect utterance may be higher than the cost of
repairing the problems that may arise through misunderstandings. For instance,
subjects are less careful about adapting their utterances to their partner when
they know they can provide feedback on his/her understanding (Schober, 1993). We
introduced the notion of ‘optimal collaborative effort’ (Dillenbourg et al,
1996) to stress that misunderstanding should not be viewed as something to be
avoided (if this was possible), but as an opportunity to engage into
verbalization, explanation, negotiation, and so forth. This issue is related to
the global argument regarding cognitive load in learning activities, especially
in discovery learning environments: there is no learning without some cognitive
load, but overload may hinder learning (Paas, Renkl \& Sweller, 2003). In the
context of collaborative learning, we understand the cognitive load induced by
mutual modeling as part of Schwartz’s (1995) notion of effort towards a shared
understanding. For instance, CSCL researchers expanded the use of 'collaboration
scripts' (Aronson,?). A script is a pedagogical method that collaborative
learning activities in order to foster the emergence of productive interactions
such as argumentation, explanation or conflict. Conflict-resolution scripts such
as the ArgueGraph (Dillenbourg \& Hong, 2008) forms pair of students with
opposite opinions, which increases the difficulty of consensus building,
requiring more justifications, more negotiation, more load. Similarly, JIGSAW
scripts provide peers with different but complementary knowledge for augmenting
(reasonably) the effort group members have to engage to reach a shared solution.

@@

    \item $Model(A,B, "urgent")$: What does A about know B's understanding of A's
        last utterance: does ‘urgent’ means now, ASAP and ‘not too lat’ ?

@@

Initial common grounds are also initiated by co-presence: they include events to
which A and B attended together~\cite{clark2002definite} in the physical space
or in our cultural space (\eg ‘09-11’). Co-presence means that we can refer to
a shared objects and event,-s it does of course not imply that we give them the
same meaning. Namely, a shared screen does not mean a shared
understanding~\cite{dillenbourg2006sharing}.

@@

\begin{equation} \label{eq:probrepair}
p(\alpha_{t_1}(X,T) / M^{\circ}_{t_2}(A,B,X) \nearrow) \approx \frac{(M^{\circ}_{min}(A,B,X,T) -
M^{\circ}_{t_1}(A,B,X))}{cost (\alpha)}
\end{equation}

@@

\subsubsection{Studies}

@@

We could go further by comparing self- versus other modeling $\Delta_4$ in
Fig.~\ref{mm_rectangle}) as an indication of metacognitive skills. We could also
see if modeling skills depend upon what aspects are being modeled (X or Y),
which would explain vertical differences ($\Delta_5$ in
Fig.~\ref{mm_rectangle}). We do not further develop these questions because we
don’t have the necessary data in our studies.

@@

@@

\subsection{Study 1: Effect of an awareness tool on \Model{A}{B}{x} in a virtual
game}

\paragraph{Questions}

We studied the impact of an awareness tool on group performance and mutual
modeling. (Nova, Wehrle, Goslin, Bourquin \& Dillenbourg (2006). The availability
of an awareness tool was our independent variable. In previous studies, we
replayed a video of the game to subjects who surprised us by their ability to
remember former states of their mutual model: "I did that because I thought that
you would do that". Hence, this experiment focused the representation of each
other's action plans. During the game, we asked them to anticipate the next
action of their partner as well as to announce their own action. It is clear
asking them a question about their partner increases their attention for the
rest of the game.

\paragraph{Experimental setting}

SpaceMiners is a 3D game that involves two players collecting minerals located
in asteroids. To do so, they shoot drones through the space after choosing their
initial direction and speed. Once launched, the trajectory of drones is only
influenced by the gravity of planets and by trajectory modification tools.
During the experiment, the teams were confronted with three increasingly complex
situations. The experiment was 2 hours long: a 30 minutes tutorial and 3 levels
of 30 minutes. Players uses using a regular joystick and communicated with each
other through an audio channel.

The independent variable was the availability of an awareness tool that shows to
player A  the location and gaze direction of player B. In the awareness
condition, players could switch to the 'scout mode' where they could view what
their partner was looking at. We hypothesize that this would enable subjects to
more accurately infer their teammate's intentions. Each player sat in front of a
distinct computer located in different rooms.

\paragraph{Subjects}

@@

\begin{figure*}
        \centering
        \includegraphics[width=0.9\textwidth]{image4.png}
        \caption{Snapshot of the SpaceMiner Game}
        \label{spaceminer}
\end{figure*}

\paragraph{Variables}

@@

In order to evaluate m°(A,B) during the task, we used two questionnaires (Figure
4) that were displayed during each of the three games, as a transparent layer
appearing over the game display. The first questionnaire concerned the player’s
intended actions. The second questionnaire asked each player about what he
thought her or his partner was intending to do. Some answers were identical in
both questionnaires (like “adjusting a shot”) while others were reversed ("to
guide my partner" versus "to guide me"). We chose this subjective measure of
accuracy (∆ M(A,B), M(B,B)) rather than an objective measure (i.e. model (A,B)
is compared to B's next action) because some the activities proposed by the
questionnaire were not observable by the environment (e.g. establishing a
strategy). We counted M°(A,B,activity,gamei) as the number of common answers
between questionnaires M(A,B) and M(B,B) in each game and computed the average
value a across the 3 games.

\begin{figure}[ht!]
        \centering
        \includegraphics[width=\columnwidth]{image5.png}
        \caption{M(A,A) and M(A,B) questionnaires in SpaceMiners (translated
        from French)}
        \label{study1:questionnaires}
\end{figure}

\paragraph{Results}

Grounding criterion. The grounding criterion was high: the correlation between
M°(A,B) and task performance was 0.42, p = 0.05. Pairs with an accurate mutual
model reached higher scores. A regression analysis confirmed the positive and
significant relation between group performance and mutual modelling accuracy (β
=54, p = .02).

Study-specific questions. The awareness tool permitted higher group performance,
but it did not improve the accuracy of the mutual model. Since teams were free
to use the awareness tool or not (the 'scout' mode), we performed a post-hoc
split of players depending on how much time they used it. The split point was
the mean of time spent in the scout mode and it led to the constitution of two
groups made up of 12 individuals “short time in scout mode” and 24 individuals
“long time in scout mode”. A two-way analysis of variance conducted on these
contrasted groups revealed that pairs in the awareness condition who spent more
time in the scout mode reached higher levels of M°(A,B)(F = 8.02, p = 0.015). Of
course, a post-hoc split does not support a causal direction. An alternative
explanation could be that good modellers are more social and hence appreciate
the awareness tool.

Symmetry question. We computed intra-class correlation as described by Kenny et
al. (1998) from the answers to the cross-questionnaires. Considering all pairs
in both conditions, we found a positive and significant correlation (r = .38, p
< .05) between M°(A,B) and M°(B,A). Interestingly, this was higher in the
control group (r = 0.44) than in the experimental group (r = 0.24). Actually, ∆
(M°(A,B),M°(B,A)), i.e. the absolute differences between the models accuracy,
was not significantly different with or without the awareness tool ( F [1,13]=
0.144, p > 0.5). This result could be explained by the fact that the players
without awareness tools communicated more.

We cannot answer the triangle questions since this study was conducted with
pairs-

\paragraph{Discussion}

How to interpret a correlation of 0.38 between M°(A,B) and M°(B,A)? If it was
null, modelling would be interpreted as an individual process that varies
according to personality traits or cognitive skills: some persons tend to model
more accurately their peer while others don't care or can simply not. If the
correlation was close to 1, we could conclude that the modelling accuracy is an
emergent property of teams since there would be low probability that, by chance,
low accurate modellers are paired with accurate modellers and conversely. Our
correlation is between these two extremes but the very notion of significant
intra-class correlation tend to indicate that one may not consider peers as
independent subjects, which is an argument for the 'interaction' hypothesis
(H3).

\subsection{Study 2: Effect of an awareness tool on M°(A,B)  in a pervasive
game}

\paragraph{Questions}

This study concerns a collaborative game that occurred in the physical space
(Nova et al , 200?). We studied whether players build an accurate model of the
path followed by their partners because this path reflected their problem
solving strategy. We used an objective measure of M°(A,B): the distance between
where A believes B has been walking and where B actually went. The main
hypothesis concerned the effect of awareness tools on group performance and on
M°(A,B).

\paragraph{Experimental setting}

CatchBob is a mobile game in which groups of 3 players have to solve a joint
task. The game was played on EPFL campus. Participants had to find a virtual
object ('Bob') and "to catch it" by forming a triangle around it. Players used a
Tablet PC that displayed a map of the campus and an indication of their personal
distance from Bob. Their annotations on the map were shared with the two other
players (A could see what B and C wrote). These annotations faded out after a
few minutes to avoid covering the full display. The awareness tool displayed the
location of the two other players on the map. The availability of location
awareness tool constituted our independent variable.

\begin{figure*}[h!t]
        \centering
        \begin{subfigure}{.3\textwidth}
            \includegraphics[width=\linewidth]{image6.png}
            \caption{A's drawing A of B's path.}
        \end{subfigure}
        \begin{subfigure}{.3\textwidth}
            \includegraphics[width=\linewidth]{image7.png}
            \caption{Actual path followed by B.}
        \end{subfigure}
        \caption{Bla bla}
        \label{study2:paths}
\end{figure*}

\paragraph{Subjects}

Ninety students participated in this experiment. We chosen only students from
our University because knowing the campus geography had an impact both on group
performance but also on mutual modeling: it is hard to represent the path of
someone across some space without having a mental map of this space. We selected
groups of students who knew each other. We assigned 10 triads to each of our
three experimental conditions. Each condition was made up of approximately 25%
of women, but we did not control gender repartition within each triad.

\paragraph{Variables}

The independent variable was the awareness tool: the control condition (without
tool) and two experimental conditions: synchronous awareness (display current
position of each player) and asynchronous awareness (display current position of
each player and their recent spatial trace).  As a dependent variable, we had
the task performance which was the distance covered by the team to catch Bob and
M°(A,B). To estimate M°(A,B), we asked players to draw on paper their own path
and the path of each of their partners after the game. This enabled us to
calculate the number of errors players made while drawing the path of their
partners. We compared the path that player A attributed to B with B's real paths
recorded by the system and the same for A\&C or B\&C as depicted on Figure 5.

M°(A,B) is the sum of errors made by A about B’s paths. An error was either
drawing a place where the partner had not been or not drawing a place where
he/she had gone. One could argue that M°(A,B) is biased by the subjects' ability
to translate their trajectories souvenir into a map drawing. However, 85% of
subjects made no mistake at all when drawing their own path. We therefore
consider mistakes in their partners' path as being due to a lack of mutual
modeling accuracy instead of being due to spatial reasoning skills.

\paragraph{Results}

Grounding criterion. The correlation between M°(A,B) and the task performance
(path lengths) was low: 0.15. Using a post-hoc split on M°(A,B), there is no
significant difference between the performance of the groups with high and low
M°(A,B)  (F = 1.45, p = 0.24). Conversely, a post-hoc split of the groups
according to their performance did not show any significant differences on
M°(A,B) (F = 1.16, p = 0.29).

Study-specific questions. There was no significant difference regarding the task
performance. However, our surprise was that the absence of the awareness tool
was related to a higher M°(A,B): players better remembered their partners’ path
if they did not see their position permanently! We will not enter into the
details of these results but, to be short, teams without awareness tool made
more annotations on the map while permanent MLA has an underwhelming effect
(Nova et al., 2005).

Symmetry question. The correlation between M°(A,B)  and M°(B,A)  is positive (r
= .41) and significant (p < .01);  the more A makes errors about B, the more B
does the same.

Triangle questions. About ∆2, the correlation between M°(A,B) and M°(A,C)is
significant: r=0.30, p <.01. About ∆3, the correlation between M°(A,C) and
M°(B,C) is significant: r=0.43, p <.001.

\paragraph{Discussion}

This positive result regarding to the symmetry question is even more surprising
than the previous one: despite the high heterogeneity of spatial skills among
adults (Liben et al, 1981), a significant correlation could indicate that mutual
modeling relies more on group processes than on individual processes. The
significant results on the triangle questions probably go in the same direction:
either the modeler or the modeled both contribute to accurate modeling, or that
the quality of teamwork has a global effect. In this experiment, since team
members did not interact intensively during the task, the factor that
facilitates accurate mutual modeling is probably less the quality of verbal
interactions that the group strategy: a clear strategy facilitates memorizing
one’s partner path.

\subsection{Study 3:  Effect of media richness on M°(A,B) in argumentation}

\paragraph{Questions}

The aim of this unpublished study was to evaluate the effect of media richness
on M°(A,B,emotions). The hypothesis was that video communication would lead to a
better M°(A,B) than audio only since emotions often impact on facial
expressions.

@@

\paragraph{Experimental settings}

>>>>

\usepackage{amsmath}

@@

\usepackage{hyperref}
\usepackage{subcaption}

\usepackage[draft, footnote, margin=false]{fixme}

@@

\newcommand{\M}[3]{{\mathcal{M}(#1, #2, #3)}}
\newcommand{\model}[3]{{$\mathcal{M}(#1, #2, #3)$}}
% generic mutual model (gmodel) 
\newcommand{\gmodel}[2]{{$\mathcal{M}(#1, #2)$}}
\newcommand{\refmodel}[2]{{$\mathcal{M}(#1, #2)$}}
\newcommand{\Model}[3]{{$\mathcal{M}^{\circ}(#1, #2, #3)$}}
\newcommand{\gModel}[2]{{$\mathcal{M}^{\circ}(#1, #2)$}}
\newcommand{\Mdeg}[3]{{\mathcal{M}^{\circ}(#1, #2, #3)}}
\newcommand{\gMdeg}[2]{{\mathcal{M}^{\circ}(#1, #2)}}

@@

From its very beginning, CSCL research has been following Roschelle \& Teasley's
(1986) suggestion that collaborative learning has something to do with the
process of constructing and maintaining a shared understanding of the task at
hand. Building a shared/mutual understanding refers to the upper class of
collaborative learning situations, those in which students should build upon
each other's understanding to refine their own understanding. What is expected
to produce learning is not the mere fact that two students build the same
understanding but the cognitive effort they have to engage to build this shared
understanding (Schwartz, 1995). This effort can be observed by the frequency of
rich interactions, \ie interactions whose occurrence has been related to
learning: (self-) explanations in cognitive science (Chi ; Webb), conflict
resolution in socio-cognitive theories (Doise), mutual regulation (Blaye,) in a
Vygostkian perspective, etc. The construction of a shared understanding has been
investigated for several years in psycholinguistics, under the  notion of
“grounding” (Clark \&Wilkes-Gibbs, 1986). However, the relevance of grounding
mechanisms for explaining learning outcomes has been questioned. The monitoring
and repair of mis-understanding explains for instance referential failures in
short dialogue episodes but does hardly predict conceptual change over longer
sessions (Dillenbourg \& Traum, 2006). The cumulative effect of grounding
episodes can probably be better understood from a socio-cultural perspective:
"collaborative learning is associated with the increased cognitive-interactional
effort involved in the transition from learning to understand each other to
learning to understand the meanings of the semiotic tools that constitute the
mediators of interpersonal interaction" (Baker, Hansen, Joiner \& Traum, 1999).
Several scholars suggest that CSCL research should go deeper in the
understanding of how partners engage into shared meaning making (Stahl, 2006) or
'intersubjective' meaning making (Suthers (2006).

Paradoxically, while Clark's theory is somewhat too linguistic from a conceptual
change viewpoint, it is criticized at the same time as being too cognitivist by
some psycholinguists, \ie as overestimating the amount shared knowledge and
mutual representations actually necessary to conduct a dialogue. The fundamental
issue, as old as philosophy, is the degree of coupling between the different
levels of dialogue, mostly between the lexical / syntactical level and the
deeper semantic levels. Pickering \& Garrod (2004) argue that the mutual
understanding starts mostly with a 'local alignment' at the level of the
linguistic representations, due to priming mechanisms, and that this superficial
alignment may – in some cases- lead to a 'global alignment' of the semantic
level.  For these authors, the convergence in dialogue, and even the repair of
some misunderstandings, is explained by this mimetic behavior more than by a
monitoring of each other's knowledge: "…interlocutors do not need to monitor and
develop full common ground as a regular, constant part of routine conversation,
as it would be unnecessary and far too costly. Establishment of full common
ground is, we argue, a specialized and non-automatic process that is used
primarily in times of difficulty (when radical misalignment becomes apparent)."
(p. 179). This view is actually not incompatible with Clark's grounding
criterion (REF): the degree of shared understanding that peers need to reach
depends upon the task they perform. For instance, a dialogue between two
surgeons might rely on superficial alignment if they talk about their friends
but has to guarantee accurate common grounds when talking about which
intervention will be conducted in which way on which patient.  Some CSCL
scholars stressed the 'illusion of shared understanding' (JACCO?), \ie the
potential decoupling between linguistic alignment and actual shared meanings.

@@

Deep grounding or shared meaning making requires some cognitive load. For Clark
\& Wilkes-Gibbs (1986), what is important is not the individual effort made by
the receiver of a communicative act, but the overall least collaborative effort.
The cost of producing a perfect utterance may be higher than the cost of
repairing the problems that may arise through misunderstandings. For instance,
subjects are less careful about adapting their utterances to their partner when
they know they can provide feedback on his/her understanding (Schober, 1993). We
introduced the notion of ‘optimal collaborative effort' (Dillenbourg et al,
1996) to stress that misunderstanding should not be viewed as something to be
avoided (if this was possible), but as an opportunity to engage into
verbalization, explanation, negotiation, and so forth. This issue is related to
the global argument regarding cognitive load in learning activities, especially
in discovery learning environments: there is no learning without some cognitive
load, but overload may hinder learning (Paas, Renkl \& Sweller, 2003). In the
context of collaborative learning, we understand the cognitive load induced by
mutual modeling as part of Schwartz's (1995) notion of effort towards a shared
understanding. For instance, CSCL researchers expanded the use of 'collaboration
scripts' (Aronson,?). A script is a pedagogical method that collaborative
learning activities in order to foster the emergence of productive interactions
such as argumentation, explanation or conflict. Conflict-resolution scripts such
as the ArgueGraph (Dillenbourg \& Hong, 2008) forms pair of students with
opposite opinions, which increases the difficulty of consensus building,
requiring more justifications, more negotiation, more load. Similarly, JIGSAW
scripts provide peers with different but complementary knowledge for augmenting
(reasonably) the effort group members have to engage to reach a shared solution.

@@

    \item $Model(A,B, "urgent")$: What does A about know B's understanding of A's
        last utterance: does ‘urgent' means now, ASAP and ‘not too lat' ?

@@

Initial common grounds are also initiated by co-presence: they include events to
which A and B attended together~\cite{clark2002definite} in the physical space
or in our cultural space (\eg ‘09-11'). Co-presence means that we can refer to
a shared objects and event,-s it does of course not imply that we give them the
same meaning. Namely, a shared screen does not mean a shared
understanding~\cite{dillenbourg2006sharing}.

@@

\begin{multline} \label{eq:probrepair}
p(\alpha_{t_1}(X,T) / M^{\circ}_{t_2}(A,B,X) \nearrow) \approx \\
\frac{(M^{\circ}_{min}(A,B,X,T) - M^{\circ}_{t_1}(A,B,X))}{cost (\alpha)}
\end{multline}

@@

\subsubsection*{Studies}

@@

We could go further by comparing self- versus other modeling $\Delta_4$ in
Fig.~\ref{mm_rectangle}) as an indication of metacognitive skills. We could also
see if modeling skills depend upon what aspects are being modeled (X or Y),
which would explain vertical differences ($\Delta_5$ in
Fig.~\ref{mm_rectangle}). We do not further develop these questions because we
don't have the necessary data in our studies.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Study 1: Effect of an awareness tool on \gModel{A}{B} in a virtual
game}

\subsubsection*{Questions}

We studied the impact of an awareness tool on group performance and mutual
modeling~\cite{nova2007collaboration}. The availability
of an awareness tool was our independent variable. In previous studies, we
replayed a video of the game to subjects who surprised us by their ability to
remember former states of their mutual model: ``I did that because I thought that
you would do that''. Hence, this experiment focused the representation of each
other's action plans. During the game, we asked them to anticipate the next
action of their partner as well as to announce their own action. It is clear
asking them a question about their partner increases their attention for the
rest of the game.

\subsubsection*{Experimental setting}

SpaceMiners is a 3D game that involves two players collecting minerals located
in asteroids (Fig.~\ref{study1:spaceminer}). To do so, they shoot drones through the space after choosing their
initial direction and speed. Once launched, the trajectory of drones is only
influenced by the gravity of planets and by trajectory modification tools.
During the experiment, the teams were confronted with three increasingly complex
situations. The experiment was 2 hours long: a 30 minutes tutorial and 3 levels
of 30 minutes. Players uses using a regular joystick and communicated with each
other through an audio channel.

The independent variable was the availability of an awareness tool that shows to
player A  the location and gaze direction of player B. In the awareness
condition, players could switch to the \emph{scout mode} where they could view what
their partner was looking at. We hypothesize that this would enable subjects to
more accurately infer their teammate's intentions. Each player sat in front of a
distinct computer located in different rooms.

\subsubsection*{Subjects}

@@

\begin{figure*}
        \centering
        \includegraphics[width=0.9\textwidth]{image4.png}
        \caption{Snapshot of the SpaceMiner Game}
        \label{study1:spaceminer}
\end{figure*}

\subsubsection*{Variables}

@@

In order to evaluate \gModel{A}{B} during the task, we used two questionnaires
(Fig.~\ref{study1:questionnaires}) that were displayed during each of the three
games, as a transparent layer appearing over the game display. The first
questionnaire concerned the player's intended actions. The second questionnaire
asked each player about what he thought her or his partner was intending to do.
Some answers were identical in both questionnaires (like ``adjusting a shot'')
while others were reversed (``to guide my partner'' versus ``to guide me''). We
chose this subjective measure of accuracy ($\Delta(\M{A}{B}{x}, \M{B}{B}{x})$)
rather than an objective measure (\ie the model \gmodel{A}{B} is compared to B's
next action) because some the activities proposed by the questionnaire were not
observable by the environment (\eg establishing a strategy). We counted
\Model{A}{B}{activity_{game_{i}}}) as the number of common answers between
questionnaires \gmodel{A}{B} and \gmodel{B}{B} in each game and computed the
average value a across the 3 games.

\begin{figure}[ht!]
        \centering
        \includegraphics[width=\columnwidth]{image5.png}
        \caption{\gmodel{A}{A} and \gmodel{A}{B} questionnaires in SpaceMiners
        (translated from French).}

        \label{study1:questionnaires}
\end{figure}

\subsubsection*{Results}

\paragraph{Grounding criterion} The grounding criterion was high: the
correlation between \gModel{A}{B} and task performance was 0.42, $p = 0.05$.
Pairs with an accurate mutual model reached higher scores. A regression analysis
confirmed the positive and significant relation between group performance and
mutual modelling accuracy ($\beta=54, p = .02$).

\paragraph{Study-specific questions} The awareness tool permitted higher group
performance, but it did not improve the accuracy of the mutual model. Since
teams were free to use the awareness tool or not (the \emph{scout} mode), we
performed a post-hoc split of players depending on how much time they used it.
The split point was the mean of time spent in the scout mode and it led to the
constitution of two groups made up of 12 individuals ``short time in scout
mode'' and 24 individuals ``long time in scout mode''. A two-way analysis of
variance conducted on these contrasted groups revealed that pairs in the
awareness condition who spent more time in the scout mode reached higher levels
of \gModel{A}{B}($F = 8.02, p = 0.015$). Of course, a \textit{post-hoc} split
does not support a causal direction. An alternative explanation could be that
good modellers are more social and hence appreciate the awareness tool.

\paragraph{Symmetry question} We computed intra-class correlation as described
by~\cite{kenny1998data} from the answers to the cross-questionnaires.
Considering all pairs in both conditions, we found a positive and significant
correlation ($r = .38, p < .05$) between \gModel{A}{B} and \gModel{B}{A}.
Interestingly, this was higher in the control group ($r = 0.44$) than in the
experimental group ($r = 0.24$). Actually, $\Delta(\gMdeg{A}{B},\gMdeg{B}{A})$,
\ie the absolute differences between the models accuracy, was not significantly
different with or without the awareness tool ($F [1,13]= 0.144, p > 0.5$). This
result could be explained by the fact that the players without awareness tools
communicated more.

\paragraph{Triangle questions} We cannot answer the triangle questions since
this study was conducted with dyads.

\subsubsection*{Discussion}

How to interpret a correlation of 0.38 between \gModel{A}{B} and \gModel{B}{A}?
If it was null, modelling would be interpreted as an individual process that
varies according to personality traits or cognitive skills: some persons tend to
model more accurately their peer while others don't care or can simply not. If
the correlation was close to 1, we could conclude that the modelling accuracy is
an emergent property of teams since there would be low probability that, by
chance, low accurate modellers are paired with accurate modellers and
conversely. Our correlation is between these two extremes but the very notion of
significant intra-class correlation tend to indicate that one may not consider
peers as independent subjects, which is an argument for the \emph{interaction}
hypothesis $\mathcal{H}_{3}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Study 2: Effect of an awareness tool on \gModel{A}{B}  in a pervasive
game}

\subsubsection*{Questions}

This study concerns a collaborative game that occurred in the physical
space~\cite{nova2006underwhelming}. We studied whether players build an accurate
model of the path followed by their partners because this path reflected their
problem solving strategy. We used an objective measure of \gModel{A}{B}: the
distance between where A believes B has been walking and where B actually went.
The main hypothesis concerned the effect of awareness tools on group performance
and on \gModel{A}{B}.

\subsubsection*{Experimental setting}

{\sc CatchBob} is a mobile game in which groups of 3 players have to solve a
joint task. The game was played on EPFL campus. Participants had to find a
virtual object (\emph{Bob}) and ``to catch it'' by forming a triangle around it.
Players used a Tablet PC that displayed a map of the campus and an indication of
their personal distance to Bob. Their annotations on the map were shared with
the two other players (A could see what B and C wrote). These annotations faded
out after a few minutes to avoid covering the full display. The awareness tool
displayed as well the location of the two other players on the map. The
availability of location awareness tool constituted our independent variable.

\begin{figure*}[h!t]
        \centering
        \begin{subfigure}{.3\textwidth}
            \includegraphics[width=\linewidth]{image6.png}
            \caption{A's drawing of B's path.}
        \end{subfigure}
        \begin{subfigure}{.3\textwidth}
            \includegraphics[width=\linewidth]{image7.png}
            \caption{Actual path followed by B.}
        \end{subfigure}
        \caption{Reported and actual path of one of the player, during the {\sc
        CatchBob} game.}
        \label{study2:paths}
\end{figure*}

\subsubsection*{Subjects}

Ninety students participated in this experiment. We chosen only students from
our university because knowing the campus geography had an impact both on group
performance but also on mutual modeling: it is hard to represent the path of
someone across some space without having a mental map of this space. We selected
groups of students who knew each other. We assigned 10 triads to each of our
three experimental conditions. Each condition was made up of approximately 25\%
of women, but we did not control gender repartition within each triad.

\subsubsection*{Variables}

The independent variable was the awareness tool: the control condition (without
tool) and two experimental conditions: synchronous awareness (display current
position of each player) and asynchronous awareness (display current position of
each player and their recent spatial trace).  As a dependent variable, we had
the task performance which was the distance covered by the team to catch Bob and
\gModel{A}{B}. To estimate \gModel{A}{B}, we asked players to draw on paper
their own path and the path of each of their partners after the game. This
enabled us to calculate the number of errors players made while drawing the path
of their partners. We compared the path that player A attributed to B with B's
real path recorded by the system and the same for A \& C and B \& C as depicted
on Fig.~\ref{study2:paths}.

\gModel{A}{B} is the sum of errors made by A about B's paths. An error was
either drawing a place where the partner had not been or not drawing a place
where he/she had gone. One could argue that \gModel{A}{B} is biased by the
subjects' ability to translate their trajectories souvenir into a map drawing.
However, 85\% of subjects made no mistake at all when drawing their own path. We
therefore consider mistakes in their partners' path as being due to a lack of
mutual modeling accuracy instead of being due to spatial reasoning skills.

\subsubsection*{Results}

\paragraph{Grounding criterion} The correlation between \gModel{A}{B} and the
task performance (path lengths) was low: 0.15. Using a \emph{post-hoc} split on
\gModel{A}{B}, there is no significant difference between the performance of the
groups with high and low \gModel{A}{B}  ($F = 1.45, p = 0.24$). Conversely, a
\emph{post-hoc} split of the groups according to their performance did not show
any significant differences on \gModel{A}{B} ($F = 1.16, p = 0.29$).

\paragraph{Study-specific questions} There was no significant difference
regarding the task performance. However, our surprise was that the absence of
the awareness tool was related to a higher \gModel{A}{B}: players better
remembered their partners' path if they did not see their position permanently!
We will not enter into the details of these results but, to be short, teams
without awareness tool made more annotations on the map while permanent MLA has
an underwhelming effect~\cite{nova2005location}.

\paragraph{Symmetry question} The correlation between \gModel{A}{B}  and
\gModel{B}{A}  is positive ($r = .41$) and significant ($p < .01$);  the more A
makes errors about B, the more B does the same.

\paragraph{Triangle questions} Regarding $\Delta_2$, the correlation between
\gModel{A}{B} and \gModel{A}{C} is significant: $r=0.30, p <.01$. Concerning
$\Delta_3$, the correlation between \gModel{A}{C} and \gModel{B}{C} is
significant as well: $r=0.43, p <.001$.

\subsubsection*{Discussion}

This positive result regarding to the symmetry question is even more surprising
than the previous one: despite the high heterogeneity of spatial skills among
adults~\cite{liben1981spatial}, a significant correlation could indicate that
mutual modeling relies more on group processes than on individual processes. The
significant results on the triangle questions probably go in the same direction:
either the modeler or the modeled both contribute to accurate modeling, or that
the quality of teamwork has a global effect. In this experiment, since team
members did not interact intensively during the task, the factor that
facilitates accurate mutual modeling is probably less the quality of verbal
interactions that the group strategy: a clear strategy facilitates memorizing
one's partner path.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Study 3:  Effect of media richness on \gModel{A}{B} in argumentation}

\subsubsection*{Questions}

The aim of this unpublished study was to evaluate the effect of media richness
on \Model{A}{B}{emotions}. The hypothesis was that video communication would
lead to a better \gModel{A}{B} than audio only since emotions often impact on
facial expressions.

@@

\subsubsection*{Experimental settings}

Triads had to address an emotional societal debate: authorizing or not adoption
by homosexual couples. They worked on-line and had to structure their
argumentation with a shared concept map tool {\sc TeamWave} as illustrated in
Fig.~\ref{study3:argumentation_graph}. Ten groups had only an audio connection
while ten groups had audio plus video. The video communication was provided by a
web cam and the software {\sc IVisit}. For the audio link, we used microphones,
headsets and a software called {\sc BattleCom}. In the audio+video condition,
the screen was divided in three sections. The main part was devoted to the
concept map window (24x24cm). In the remaining part appeared the image of the
two peers (8.5x8.5cm each). In the audio condition, this video zone was left
empty so that the size of the concept map was equal in each condition. The
subjects were located in the same room, separated by mobile walls. Despite their
headsets, it is not impossible that they could hear non-verbal audio cues (\eg
tapping the sole with feet).  The task last in average 61 minutes.

\subsubsection*{Subjects}
